---
title: "R Notebook"
output: html_notebook
---

Data Exploration
-----------------------------

Open libraries
-----------------------------

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(data.table)
library(forcats)
library(scales)
```

 Set the file path
--------------------------

```{r}
dirpath <- "C:/Users/Eric/Documents/Eric/Pro/Transition/Formation/CEPE ENSAE ENSAI Certificat Data Scientist/INTENSIVE/PROJETS/KickClub Project"
dirpath2 <- "Lending Club/LC0715"
filename <- "loan.csv"
filename <- "LC0715.RDS"
savefile <- "LC0715.RDS"
my_file <- paste(dirpath, dirpath2, filename, sep = "/")
my_file
```

Load data
--------------------------

Check the file's format before retrieving the original data
Retrieve the original data set

```{r}
scan(file = my_file, what = "string", nmax = 10)
my_data <- read.csv(my_file, sep = ",")
```

Or alternatively, for a faster loading use fread from data.table package
beware, having 'factor' or 'character' variables is very different!

```{r}
my_data <- fread(my_file, stringsAsFactors = TRUE)
```

Save the data in RDS format
```{r}
saveRDS(object = my_data, file = paste(dirpath, dirpath2, savefile, sep = "/"))
```

```{r}
my_data <- readRDS(my_file)
```


Quick data summary
-------------------------

Basics

```{r}
str(my_data)
summary(my_data)
glimpse(my_data)
names(my_data)
which_class <- sapply(my_data, class)
```

Note: which_class definition coded below is much simpler than the alternative:
  " which_class <- sapply(my_data, function(col){class(col)}) ""
  

Get the variables' classes
Different classes available in the dataset

```{r}
variables_types <- unique(sapply(my_data, class))
```

Calculate the number of variables belonging to each class 

```{r}
table(sapply(my_data, class))
```

Calculate the percentage of NAs for each variable

```{r}
perc_na <- round(sapply(my_data, function(x) sum(is.na(x)/length(x))),4)
perc_na
```


Draw a graph of percentage NAs 'perc_na' with variable names labels and factor classes color code
------------ Method #1 ------------------

1) Get the names of variables in a vector

```{r}
var_name <- names(my_data) # note: all.equal(names(my_data), names(perc_na)) = TRUE
```

2) we need a data.frame to apply ggplot2

```{r}
var_summary <- as.data.frame(cbind(var_name, perc_na, which_class))
var_summary
sapply(var_summary,class) # perc_na is a 'factor'
var_summary$perc_na <- as.numeric(as.character(var_summary$perc_na))
names(var_summary)
```

3) Arrange data order before drawing the grap

```{r}
var_summary <- var_summary %>% arrange(perc_na, which_class)
```


4) Draw the graph: percentage of NAs vs. variable name, ordered by indreasing NAs % and variable class

```{r}
var_summary %>% 
  ggplot((aes(x = reorder(var_name, order(var_summary$perc_na, decreasing = FALSE)), y = perc_na, color = which_class))) + 
  geom_point(shape = 18, size = 4) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 10)) +
  labs(x = names(var_summary), title = "Proportion of NAs across variables")  # or, x = names(var_summary)[1]
```

5) clean up the environment

```{r}
rm(var_summary, var_name, perc_na)
```

------------ Method #2 -------------------

1) Create a data.frame or a list of all the different classes/types 
Create a dataframe (vector) of "is.'class'" to be used afterwards

```{r}
# is.variables_types <- c("is.integer", "is.numeric", "is.factor") # what we would like to achieve
# is.variables_types <- data.frame() # this simple empty data.frame is not enough since we may need to set the dimensions
is.variables_types <- data.frame(1:length(variables_types)) # we set the number of columns
```

Or

```{r}
is.variables_types <- list() # setting the dimension is not necessary in the case of a list
```

This loop will feed the different variables types into our data.frame (or list)

```{r}
for (i in 1:length(variables_types)) {
  is.variables_types[i] <- paste("is.",variables_types[i], sep = "")
}
rm(i) 
```

2) Variables types summary: build a summary list of all the variables according to their respective class/type
We need a list to create 3 different lists of different size

```{r}
variables_types_summary <- vector("list", length(variables_types)) # set an empty list of 'length(variables_types)' elements
names(variables_types_summary) <- variables_types # set its header
```

The loop below fills in the different elements of the list with the name of variables belonging to the different classes/types
Index the variables of type [i], to be used if is.variables_types is c("is.integer", "is.numeric", "is.factor")

```{r}
for (i in 1:length(is.variables_types)) {
  indx <- sapply(my_data, is.variables_types[[i]][1])
  variables_types_summary[[i]] <- names(my_data)[indx]  # input the variable name in the relevant class column
}
variables_types_summary
rm(is.variables_types, variables_types, indx, i)
```
 
3) Calculate different inputs

The share of NAs for each variable

```{r}
sum_na <- sapply(my_data, function(col){sum(is.na(col))})
sum_na
ratio_na <- round(sum_na/dim(my_data)[1],4)
ratio_na
```

Set var_name vecvtor 

```{r}
var_name <- names(my_data) # or names(ratio_na)
```
 
4) Create a 'descriptor' data.frame that binds the different useful vectors

```{r}
descriptor <- data.frame(cbind(var_name,which_class, sum_na, ratio_na)) # all columns are factors
```

Then, convert relevant data in numeric type, and reorganise and rearrage data

```{r}
descriptor <- descriptor %>% 
  mutate(sum_na_n = as.numeric(as.character(sum_na)), ratio_na_n = as.numeric(as.character(ratio_na))) %>% 
  select(-sum_na, -ratio_na) %>% 
  mutate(ratio_na = ratio_na_n, sum_na = sum_na_n) %>% 
  select(-ratio_na_n, -sum_na_n) %>% 
  arrange(ratio_na, which_class)
```

Check the variables classes and get rid of no longer needed vectors

```{r}
descriptor %>% sapply(class) 
rm(ratio_na, sum_na) #  

```
 
5) Draw the graph: percentage of NAs vs. variable name, ordered by indreasing NAs % and variable class

```{r}
descriptor %>% 
  ggplot(aes(x = reorder(var_name, order(ratio_na, decreasing = FALSE)), y = ratio_na, color = which_class)) + 
  geom_point(shape = 18, size = 4) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 10)) +
  labs(x = names(descriptor)[1], title = "Proportion of NAs across variables") # or, x = names(descriptor) 
```

clean up the environment

```{r}
rm(descriptor, variables_types_summary)
```

FYI: ggplot shapes infos: http://www.sthda.com/english/wiki/ggplot2-point-shapes





What are the 'factor' variables ?
--------------------------

1) Create a list of the different levels available for each 'factor' variable

```{r}
fac_lev <- sapply(my_data[ , which_class == "factor", with = FALSE], levels)
class(fac_lev)

```

2) Convert a data.table into a data.frame
If my_data is indeed a data.frame? (ie. you did load original data with fread())

```{r}
my_data_df <- as.data.frame(my_data)
dim(my_data_df[ , which_class == "factor"])
```
Note simple subsetting works well with a data.frame instead of a data.table


How many levels do 'factor' variables have? Could we simplify them?
-------------------------------------------------------------------

1) Create vectors with the nb of levels of the different factors and a vetor with the name of the factors
```{r}
fac_lev_nb <- sapply(fac_lev, length)
fac_name <- var_name[which_class == "factor"]
```
 

2) There are so many levels for given factors that you may look at them one by one
Set i, and check factors one by one

```{r}
i <- 1
  fac_name[i] # factor name
  fac_lev[[i]] # factor's levels
  length(fac_lev[[i]]) # number of levels (sometimes, there are too many)
  # plot(my_data_df %>% select(id, fac_name[i]) %>% arrange(id) %>% select(fac_name[i]) %>% table()) # this grpah is not best
  # Order the frequecy of levels in %
  as.data.frame(my_data_df %>% select(fac_name[i]) %>% table()) %>% 
    mutate(perc = round(Freq/sum(Freq), 3)) %>%
    arrange(desc(perc))
rm(i)
```

3) Focus on "loan_status" variable 
This variable will be key to assess defaulting borrowers later

```{r}
as.data.frame(my_data_df %>% select(loan_status) %>% table()) %>% 
  mutate(perc = round(Freq/sum(Freq), 3)) %>%
  arrange(desc(perc))
```


Deal with date variables : method Lubridate
--------------------------------------------

1) "issue_d" (i = 8) represents the loan issue date but is a factor : we must convert this variable in 'date' format

```{r}
my_data_df[ , "issue_d"] <- lubridate::mdy(my_data_df[ , "issue_d"])
my_data_df[1:12 , "issue_d"]
class(my_data_df[,"issue_d"]) # date
```

2) Create a Table and convert it to a data.frame that summarizes the number of loans (id) by date issue (issue_d) 

```{r}
dt_to_g <- my_data_df %>% select(issue_d) %>% table()
dt_to_g <- as.data.frame(dt_to_g) # this is a table = convert in df
str(dt_to_g) # date has been converted into factor again
colnames(dt_to_g) <- c("date","nb_of_loans") # re-name columns
head(dt_to_g)
dt_to_g[ , "date"] <- lubridate::ymd(dt_to_g[ , "date"]) # convert date variable into date format
str(dt_to_g)
```
 
3) Plot the number of loans over time (monthly)

```{r}
dt_to_g %>% ggplot(aes(x = date, y = nb_of_loans)) + 
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
  scale_x_date(breaks = date_breaks("years"), labels = date_format("%b-%Y")) + 
  labs(title = "Number of Loans issued by Month")
```

Plot the cumulative nb of loans

```{r}
dt_to_g$cum_nb_of_loans <- cumsum(dt_to_g$nb_of_loans)

dt_to_g %>% ggplot(aes(x = date, y = cum_nb_of_loans)) + 
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
  scale_x_date(breaks = date_breaks("years"), labels = date_format("%b-%Y")) + 
  labs(title = "Cumulative Number of Loans issued by Month")

 
```

Clean up
```{r}
rm(dt_to_g)
```


What are the 'numeric' variables ?
-----------------------------------

```{r}
dim(my_data_df[ , which_class == "numeric"])
unique(my_data_df[ , which_class == "numeric"] %>% names())
```

What do they look like





 